---
title: "Midterm Project - Hawaiin Survey Data"
author: "Jacob Burke"
date: "19/10/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(knitr.table.format = "html")
library(tidyverse)
library(asciiSetupReader)
library(dplyr)
library(knitr)
library(kableExtra)
library(magrittr)
library(tidyverse)
library(esquisse)  
```

## Data Set

- For this project, we are using the Hawaiin Health Survey data. 
- Our 'modeling team' wanted a tidy data set to use for logistic regression, to analyze the observed risks 
  of being diagnosed with diabetes, in relation to a number of different predictors 

## Steps Taken

The steps taken to prepare this data for analysis was as follows:  
1. Access raw survey data file  
2. Ensure data was of sufficient volume and uniqueness  
3. Extract variables of interest (diabetes and predictors)  
4. Clean these multiple data subsets  
5. Bring subsets together to form new Tidy data set  
6. EDA on Tidy data, preparing for further statisical analysis & modeling 

## Data Volume & Uniqueness Check 

- Once the raw data was extracted, counts of NA values for each 
  variable, as well as uniqueness of responses were needed to determine 
  what variables were sufficient for analysis.   
- Data generated (preview subset): 

```{r, echo = F}

hawaii <- read_ascii_setup("samadult.dat", "SAMADULT.sps")

hawaii <- as_tibble(hawaii)

hawaiiNA <- NA

for(i in 1:488)
{
 hawaiiNA <- rbind(hawaiiNA, sum(is.na(hawaii[ , i])))
}

hawaii_unique <- NA 

for(i in 1:488){
  hawaii_unique <- rbind(hawaii_unique, length(t(unique(hawaii[,i]))))
}


hawaii_answers <- cbind(hawaiiNA, hawaii_unique)
colnames(hawaii_answers) <- c("NACount", "UniqueAnswers")

## frequency table 
hawaii_answers <- hawaii_answers[2:489, ]

hawaii_answers <- t(hawaii_answers)
```

```{r answers, echo = F}
print(head(t(hawaii_answers)))

```

## Extracting Variables 

Variables of interest for this analysis were as follows:  
  
- Outcome: Diabetes (binary (0,1))  
- Predictor: Height(in) (categorical)  
- Predictor: Weight(lbs) (categorical)  
- Predictor: BMI (categorical)  
- Predictor: Alcohol Consumption (categorical)  
- Predictor: Sick Days Taken Annually (categorical)  
- Predictor: Sleep Hours per Night (categorical)  
- Predictor: Joint Pain (binary(1,2))

## Cleaning the data 

The different aspects required to clean these extracted variables were as follows:  
  
- Recode missing values from each variable (eg: 997, 998, 999, etc.) to NA's 
- Recode the diabetes variable categorization to a binary set of (0,1)
  (assumptions that were made are documented in the report)  
- Settle any other abnormalities in the predictors variable data (documented in report)  
- Rename column variable names for easier interpretation  
- Remove all final observation rows containing any NA values 

## Tidy Data

From here, we generated a new clean data set as shown: (header preview)  

```{r, echo = F}
diabetes <- hawaii[, 47]

colnames(diabetes) <- "Diabetes"

diabetes[diabetes == 2] <- 0
diabetes[diabetes == 3] <- 1
diabetes[diabetes == 7] <- NA
diabetes[diabetes == 9] <- 0

## Physical attributes 

physical <- hawaii[, 315:317]

physical <- as_tibble(physical)

## Renaming columns 

colnames(physical) <- c("Height_(in)", "Weight_(lbs)", "BMI")

## Setting NA values to 96, 97, 98, 99 values entered in height (these were either refused answers, or not certain on height)

physical$`Height_(in)`[physical$`Height_(in)` == 96 | physical$`Height_(in)` ==97 
                      | physical$`Height_(in)`== 98
                      | physical$`Height_(in)` ==99] <- NA

## Setting unanswered Weight values to NA 

physical$`Weight_(lbs)`[physical$`Weight_(lbs)` == 996 | physical$`Weight_(lbs)` ==997 
                      | physical$`Weight_(lbs)` == 998
                      | physical$`Weight_(lbs)` ==999] <- NA

## Setting unanswered BMI Values to NA 

physical$BMI[physical$BMI == 5] <- NA

## 1) Alcohol 

alcohol <- hawaii[,313]

## setting unknown drinking category to NA 

alcohol[alcohol == 10] <- NA
alcohol <- as_tibble(alcohol)
colnames(alcohol) <- "Alcohol_Consum"

## 2) General Health 

health <- hawaii[,224:226]

## Looking at the health data, I've decided to use the "number of work days lost in the past 12 months" as the best overall proxy for general health. 

health <- health[,1]
colnames(health) <- ("Sick_Days_General")
health[health == 999] <- NA 

## 3) Sleep 

sleep <- hawaii[,472:476]
## Looking at the sleep data, "Hours of sleep" will be the survey question I'm choosing as my sleeping habits proxy for these individuals


sleep <- sleep[,1]
colnames(sleep) <- ("Hours_Sleep_Night")
sleep[sleep == 97 | sleep == 98 | sleep == 99] <- NA

## 4) Joint pain 

joint <- hawaii[, 58:78]

## Looking at this section, "joint pain last 30 days" will be the variable subset I'll use as my joint pain proxy 

joint <- joint[,1]

## cleaning values that were not answered, or did not know 

joint[joint == 7 | joint == 9] <- NA
colnames(joint) <- "Joint_Pain"

data <- cbind(diabetes, physical, alcohol, health, sleep, joint)

## To remove final NA values 

data <- na.omit(data)

```

```{r, echo = F}
## Tidy data (preview)

print(head(data))

```

## EDA

Once our tidy data set was generated, we then began to run EDA on diabetes data, to try and see what patterns or certain observations would arise.   

We chose to run multiple histograms on a number of the predictor variables (excluding physical attributes height
weight), to look at the distributions of survey response values for each. Since each predictor variable organized their data in a categorical sense, we will also show corresponding categorization tables to each visual. 

## Alcohol

```{r, echo = F}
Number <- (1:9)

Category <- c("Lifetime abstainer", "Former infrequent", 
      "Former regular", 'Former, unknown frequency', 'Current infrequent', 'Current light',  'Current moderate', 'Current heavier', 'Current drinker, frequency/level unknown')

alcohol_cat <- cbind(Category,Number)

kable(alcohol_cat, format = "html", booktabs=TRUE, digits = 2, caption = 
        "Alcohol Categorization") %>% kable_styling(font_size = 15)

```
## Alcohol {.smaller}

```{r}
ggplot() + geom_histogram(aes(x = data$Alcohol_Consum), binwidth = 1, color = 'black', fill = 'blue') + labs(x = "Alcohol Consumption Categorized (1-9)", y = "Frequency", title = "Alcohol Consumption")
```
- Here we essentially have a split between individuals who are former drinkers or abstainers, and a larger subset who are current drinkers, with most identifying on the 'light' side, while still a significant amount on the moderate-heavy side.

## BMI

```{r}

Number <- (1:4)

Category <- c('Underwieght', 'Healthy Weight', 'Overweight', 'Obese')

BMI_cat <- cbind(Category,Number)

kable(BMI_cat, digits = 2, format = "html", booktabs=TRUE, caption = "BMI Categorization") %>% kable_styling()

```

## BMI {.smaller}

```{r}

ggplot() + geom_histogram(aes(x = data$BMI), binwidth = 1, color = 'black', fill = 'red') + labs(x = "BMI", y = "Frequency", title = "BMI")

```

- From here we can see that there is a significant skew of these individuals either overweight, or obese. This could suggest we're looking at quite a few individuals in this data set with less than optimal health.

## Count of Sick Days Taken per Year

```{r}
## Health

Number <- (1:2)
Number[1] <- "0-18"
Number[2] <- "19"

Category <- c('0-18 days', '19 or more days')

Health_cat <- cbind(Category,Number)

kable(Health_cat, digits = 2, format = "html", booktabs=TRUE, caption = "Sick Days Categorization") %>% kable_styling() 

```

## Count of Sick Days Taken per Year {.smaller}

```{r}

ggplot() + geom_histogram(aes(x = data$Sick_Days_General), binwidth = 1, color = 'black', fill = 'green') + labs(x = "Annual Number of Sick Days Taken", y = "Frequency", title = "Sick Days")
```
 - We can see that most individuals missed little to no time from work due to illnesses. There are a significant few between 2-5 sick days taken, and then a significant outlier subset of a few people who claim they missed more than 15 days. 

## Sleep Hours per Night
```{r}
Number <- (1:3)

Number[1] <- "3"
Number[2] <- "4-11"
Number[3] <- "12"

Category <- c("1-3 hours", "4-11 hours", 
      "12 or more hours")

Sleep_cat <- cbind(Category,Number)

kable(Sleep_cat, digits = 2, format = "html", booktabs=TRUE, caption = "Sleep Amounts Categorization") %>% kable_styling() 

```

## Sleep Hours per Night {.smaller}

```{r}
ggplot() + geom_histogram(aes(x = data$Hours_Sleep_Night), binwidth = 1, color = 'black', fill = 'yellow') + labs(x = "Average Hours of Sleep per Night", y = "Frequency", title = "Nightly Sleep Hours")
```
- From here, we can see that most average at just about 6 hours of sleep a night, with the majority of individuals hovering between 5-8 hours.

## Concluding/Next Steps

From here, the 'modeling team' can take this cleaned and explored data set to start analysis, with the main goal of creating a logistic regression model to analyze the risks of diabetes diagnosis.

